\pdftitle

### 2018-05-18 maxNt_v3.tex
V3

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2018-05-18T11:50:42+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\usepackage[final,%
nonatbib]{nips_2017}

@@

\newcommand*{\pdftitle}{Maximum-entropy and samples of neuronal activity:\\an alternative approach}

@@

\author{
  P.G.L. Porta Mana\\\texttt{piero.mana@ntnu.org}
  \And
  V. Rostami\\\texttt{v.rostami@fz-juelich.de}
  \AND
  E. Torre\\\texttt{torre@ibk.baug.ethz.ch}
    \And
Y. Roudi\\\texttt{yasser.roudi@ntnu.no}
  % David S.~Hippocampus\thanks{Use footnote for providing further
  %   information about author (webpage, alternative
  %   address)---\emph{not} for acknowledging funding agencies.} \\
  % Department of Computer Science\\
  % Cranberry-Lemon University\\
  % Pittsburgh, PA 15213 \\
  % \texttt{hippo@cs.cranberry-lemon.edu} \\
  %% examples of more authors
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \AND
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
}

@@

\begin{abstract}
  This note shows that the \me\ method can be applied to a sample of
  neuronal activity along two different routes: (1) apply to the sample; or
  (2) apply to a larger, unsampled neuronal population from which the
  sample is drawn, and then marginalize to the sample. The second route can
  be generalized to the case where the size of the larger population is
  unknown. These two routes give inequivalent results. Which route should
  be chosen? Some arguments are presented in favour of the second. This
  note also presents and discusses probability formulae that relate our
  states of knowledge about a population and its samples, and that may be
  useful for sampling problems in neuroscience.
  % The abstract must be limited to one paragraph.
\end{abstract}

@@

\enquote{Cooperativity} can quantified and characterized with \me\ method
in several ways. The simplest way roughly proceeds along the following
steps. Consider the recorded activity of a sample of $n$ neurons.
\begin{enumerate}
\item The activity of each neuron, a continuous signal, is divided into $T$
  time bins and binarized in intensity, and thus transformed into a
  sequence of digits \enquote{$0$}s (inactive) and \enquote{$1$}s
  \cite[\cf][]{caianiello1961,caianiello1986}.

@@

The purpose of the present work is not to assess the merits of \me\ methods
with respect to other methods. Its main purpose is to show that there is a
problem in the way the \me\ method itself, as sketched above, is applied to
the activity of the recorded neurons. We believe that this problem is at
the root of some quirks about this method that were pointed out in the
literature \cite{roudietal2009b}. This problems extends also to more
complex versions of the method, possibly except version that use
\enquote{hidden} neurons \cite{smolensky1986,hintonetal2006}. The problem
is that the recorded neurons are a \emph{sample} from a larger, unrecorded
population, but the \me\ method as applied above is treating them as
isolated from the rest of the brain. Hence, the results it provides cannot
rightfully be extrapolated. We will give a mathematical proof of this. Let
us first analyse this issue in more detail.

@@

In this paper we develop the second way of applying the \me\ method, at the
level of the larger population, and show that its results differ from the
application at the sample level. We also consider the case where the size
of the larger population is unknown.

@@

The notation and terminology in this note follow ISO and ANSI standards
\citep{iso1993,ieee1993,nist1995,iso2006,iso2006b} but for the use of the
comma \enquote{,} to denote logical conjunction. Probability notation
follows Jaynes \citep{jaynes1994_r2003}. By \enquote{probability} we mean a
degree of belief which \enquote{would be agreed by all rational men if
  there were any rational men} \cite{good1966}.

@@

The purpose of the present note was to simply to point out and show, in a
simple set-up, that the \me\ method can be applied to recorded neuronal
data in a way that accounts for the larger population from which the data
are sampled,
\eqns~\eqref{eq:app_maxent_pop}--\eqref{eq:app_maxent_pop_marg}. This
application leads to results that differ from the standard application
focused on the sample only,
\eqns~\eqref{eq:app_maxent_sample}--\eqref{eq:app_maxent_sample_constraints}.
We gave a numerical example of this difference. We have also shown how to
extend the new application when the size of the larger population is
unknown, \eqn~\eqref{eq:app_maxent_pop_marg_unknown_N}.

The latter formula, in particular, shows that the standard way of applying
\me\ 
implicitly assumes that \emph{no} larger population exists beyond the
recorded sample of neurons. One could in fact object to the application at
the population level, and say that the traditional way of applying \me,
\eqn~\eqref{eq:app_maxent_sample}, yields different results because it does
not make assumptions about the size $\yNv$ of a possibly existing larger
population. Such a state of uncertainty, however, is correctly formalized
according to the laws of probability by introducing a probability
distribution for $\yNv$, and is expressed by
\eqn~\eqref{eq:app_maxent_pop_marg_unknown_N}. This expression cannot
generally be equal to~\eqref{eq:app_maxent_sample} unless the distribution
for $\yNv$ gives unit probability to $\yNv=n$; that is, unless the sample
\emph{is} the full population, and no larger population exists or the
recorded neurons are completely isolated from it.

This hidden assumption of isolation is unreasonable. It amounts to say that
the neurons we distinguished and tracked with our recording device are a
very special, isolated set among all those that could have been recorded.
It is for this reason that we find the \me\ application at the population
level preferable. Physical models of neuronal networks usually include some
sort of external input, mimicking an embedding in a larger network. The
\me\ distribution obtained at the population level may reveal features in a
data set that were unnoticed by the standard \me\ model.

The difference in the resulting distributions between the applications at
the sample and at the population levels appears in the use of Boltzmann
machines with hidden units \cite{lerouxetal2008}, although by a different
conceptual route. It also appears in statistical mechanics: if a system is
statistically described by a \me\ Gibbs state, its subsystems cannot be
described by a Gibbs state \cite{maesetal1999}. A somewhat similar
situation also appears in the statistical description of the final state of
a non-equilibrium process starting and ending in two equilibrium states: we
can describe our knowledge about the final state either by (1) a Gibbs
distribution, calculated from the final equilibrium macrovariables, or (2)
by the distribution obtained from the Liouville evolution of the Gibbs
distribution assigned to the initial state. The two distributions differ
(even though the final \emph{physical} state is obviously exactly the same
\cite[\sect~4]{jaynes1985d_r1993}), and the second allows us to make
sharper predictions about the final physical state thanks to our knowledge
of its preceding dynamics. In this example, though, both distributions are
usually immensely sharp and practically lead to the same predictions. In
neuroscientific applications, the difference in predictions of the sample
\vs\ full-population applications can instead be very relevant.

The idea of the new application leads in fact to more questions. For
instance:
\begin{itemize}
\item Do the standard and new applications lead to different or contrasting
  conclusions about \enquote{cooperativity}, when applied to real data
  sets?
\item How to extend the new application to the \enquote{inhomogeneous} case
  \cite{schneidmanetal2006,shlensetal2006,roudietal2009b}, in which
  expectations for individual neurons or groups of neurons are constrained?
\item What is the mathematical relation between the new application and
  \me\ models with hidden neurons \cite{smolensky1986,hintonetal2006}?
\end{itemize}
Owing to space limitations we cannot investigate these questions here, but
we hope to in future work.

Finally, we would like to point out the usefulness and importance of the
probability formulae that relate our states of knowledge about a population
and its samples, presented in \sect~\ref{sec:prob_samples}. This kind of
formulae is essential in neuroscience, where we try to understand
properties of extended regions from partial observations of them. The
formulae presented here reflected a very basic, symmetric state of
ignorance. It would be useful to explore formulae more appropriate to our
real knowledge of the brain and of its networking properties.

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2018-05-18T15:43:51+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\usepackage[%final,%
nonatbib]{nips_2017}

@@

\newcommand*{\pdftitle}{Maximum-entropy and representative samples\\ of neuronal
  activity: a dilemma}

@@

\author{
  P.G.L. Porta Mana\\\texttt{piero.mana@ntnu.no}
  \And
  V. Rostami\\\texttt{vrostami@uni-koeln.de}
  \AND
  E. Torre\\\texttt{torre@ibk.baug.ethz.ch}
    \And
Y. Roudi\\\texttt{yasser.roudi@ntnu.no}
  % David S.~Hippocampus\thanks{Use footnote for providing further
  %   information about author (webpage, alternative
  %   address)---\emph{not} for acknowledging funding agencies.} \\
  % Department of Computer Science\\
  % Cranberry-Lemon University\\
  % Pittsburgh, PA 15213 \\
  % \texttt{hippo@cs.cranberry-lemon.edu} \\
  %% examples of more authors
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \AND
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
}

@@

\begin{abstract}
  The present work shows that the \me\ method can be applied to a sample of
  neuronal recordings along two different routes: (1) apply to the sample;
  or (2) apply to a larger, unsampled neuronal population from which the
  sample is drawn, and then marginalize to the sample. These two routes
  give inequivalent results. The second route can be further generalized to
  the case where the size of the larger population is unknown. Which route
  should be chosen? Some arguments are presented in favour of the second.
  This work also presents and discusses probability formulae that relate
  states of knowledge about a population and its samples, and that may be
  useful for sampling problems in neuroscience.
  % The abstract must be limited to one paragraph.
\end{abstract}

@@

\enquote{Cooperativity} can be quantified and characterized with \me\
methods in several ways. The simplest way roughly proceeds along the
following steps. Consider the recorded activity of a sample of $n$ neurons.
\begin{enumerate}
\item The activity of each neuron, a continuous signal, is divided into $T$
  time bins and binarized in intensity, and thus transformed into a
  sequence of digits \enquote{$0$}s (inactive) and \enquote{$1$}s
  \cite[\cf][]{caianiello1961,caianiello1986}.

@@

The purpose of the present work is not to assess the merits of \me\ methods
with respect to other methods. Its main purpose is to show that there is a
problem in the way the \me\ method itself, as sketched above, is applied to
the activity of the recorded neurons. We believe that this problem is at
the root of some quirks about this method that were pointed out in the
literature \cite{roudietal2009b}. This problems extends also to more
complex versions of the method, possibly except version that use
\enquote{hidden} neurons
\cite{smolensky1986,kulkarnietal2007,huang2015,dunnetal2017}. The problem is
that the recorded neurons are a \emph{sample} from a larger, unrecorded
population, but the \me\ method as applied above is treating them as
isolated from the rest of the brain. Hence, the results it provides cannot
rightfully be extrapolated. We will give a mathematical proof of this. Let
us first analyse this issue in more detail.

@@

In this work we develop the second way of applying the \me\ method, at the
level of the larger population, and show that its results differ from the
application at the sample level. We also consider the case where the size
of the larger population is unknown.

@@

The notation and terminology in the present work follow ISO and ANSI standards
\citep{iso1993,ieee1993,nist1995,iso2006,iso2006b} but for the use of the
comma \enquote{,} to denote logical conjunction. Probability notation
follows Jaynes \citep{jaynes1994_r2003}. By \enquote{probability} we mean a
degree of belief which \enquote{would be agreed by all rational men if
  there were any rational men} \cite{good1966}.

@@

The purpose of the present work was to point out and show, in a simple
set-up, that the \me\ method can be applied to recorded neuronal data in a
way that accounts for the larger population from which the data are
sampled, \eqns~\eqref{eq:app_maxent_pop}--\eqref{eq:app_maxent_pop_marg}.
This application leads to results that differ from the standard application
which only considers the sample in isolation,
\eqns~\eqref{eq:app_maxent_sample}--\eqref{eq:app_maxent_sample_constraints}.
We gave a numerical example of this difference. We have also shown how to
extend the new application when the size of the larger population is
unknown, \eqn~\eqref{eq:app_maxent_pop_marg_unknown_N}.

The latter formula, in particular, shows that the standard way of applying
\me\ 
implicitly assumes that \emph{no} larger population exists beyond the
recorded sample of neurons. One could in fact object to the application at
the population level, and say that the traditional way of applying \me,
\eqn~\eqref{eq:app_maxent_sample}, yields different results because it does
not make assumptions about the size $\yNv$ of a possibly existing larger
population. Such a state of uncertainty, however, is correctly formalized
according to the laws of probability by introducing a probability
distribution for $\yNv$, and is expressed by
\eqn~\eqref{eq:app_maxent_pop_marg_unknown_N}. This expression cannot
generally be equal to~\eqref{eq:app_maxent_sample} unless the distribution
for $\yNv$ gives unit probability to $\yNv=n$; that is, unless the sample
\emph{is} the full population, and no larger population exists.

The standard \me\ approach therefore assumes that the recorded neurons
constitute a special subnetwork, isolated from the larger network of
neurons in which it is embedded, and which was also present under the
recording device. This assumption is unrealistic. The \me\ approach at the
population level does not make such assumption and is therefore preferable.
It may reveal features in a data set that were unnoticed by the standard
\me\ approach.

% This hidden assumption of isolation is unreasonable. It amounts to say that
% the neurons we distinguished and tracked with our recording device are a
% very special, isolated set among all those that could have been recorded.
% It is for this reason that we find the \me\ application at the population
% level preferable. Physical models of neuronal networks usually include some
% sort of external input, mimicking an embedding in a larger network. The
% \me\ distribution obtained at the population level may reveal features in a
% data set that were unnoticed by the standard \me\ model.

The difference in the resulting distributions between the applications at
the sample and at the population levels appears in the use of Boltzmann
machines with hidden units \cite{lerouxetal2008}, although by a different
conceptual route. It also appears in statistical mechanics: if a system is
statistically described by a \me\ Gibbs state, its subsystems cannot be
described by a Gibbs state \cite{maesetal1999}. A somewhat similar
situation also appears in the statistical description of the final state of
a non-equilibrium process starting and ending in two equilibrium states: we
can describe our knowledge about the final state either by (1) a Gibbs
distribution, calculated from the final equilibrium macrovariables, or (2)
by the distribution obtained from the Liouville evolution of the Gibbs
distribution assigned to the initial state. The two distributions differ
(even though the final \emph{physical} state is obviously exactly the same
\cite[\sect~4]{jaynes1985d_r1993}), and the second allows us to make
sharper predictions about the final physical state thanks to our knowledge
of its preceding dynamics. In this example, though, both distributions are
usually extremely sharp and practically lead to the same predictions. In
neuroscientific applications, the difference in predictions of the sample
\vs\ full-population applications can instead be very relevant.

The idea of the new application leads in fact to more questions. For
instance:
\begin{itemize}
\item Do the standard and new applications lead to different or contrasting
  conclusions about \enquote{cooperativity}, when applied to real data
  sets?
\item How to extend the new application to the \enquote{inhomogeneous} case
  \cite{schneidmanetal2006,shlensetal2006,roudietal2009b}, in which
  expectations for individual neurons or groups of neurons are constrained?
\item What is the mathematical relation between the new application and
  \me\ models with hidden neurons
  \cite{smolensky1986,kulkarnietal2007,huang2015,dunnetal2017}?
\end{itemize}
Owing to space limitations we must leave a thorough investigation of these
questions to future work.

Finally, we would like to point out the usefulness and importance of the
probability formulae that relate our states of knowledge about a population
and its samples, presented in \sect~\ref{sec:prob_samples}. This kind of
formulae is essential in neuroscience, where we try to understand
properties of extended brain regions from partial observations. The
formulae presented here reflect a simple, symmetric state of ignorance.
More work is needed \cite[\cf][]{levinaetal2017} to extend these formulae
to account for finer knowledge of the cerebral cortex and its network
properties.
